---
title: "Gender Role Attitudes and Life Satisfaction: A Cross-National Analysis"
subtitle: "Data from the World Values Survey"
author: "Mia Yan"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

------------------------------------------------------------------------

# Project Note

This project began as a lab assignment for *Stats for Social Research*.

**To reproduce:** download `WVS_Cross-National_Wave_7_csv_v6_0.csv` from the
[World Values Survey site](https://www.worldvaluessurvey.org/WVSDocumentationWV7.jsp)
and place it in the same folder as this `.Rmd` file, then knit.

------------------------------------------------------------------------

# Introduction

## Motivation

Gender role attitudes (beliefs about whether women's primary place is in the home
or the workforce) vary widely across societies and are frequently studied as both
a cause and a consequence of women's economic and social status. A separate literature
links subjective well-being to the "fit" between an individual's attitudes and the
norms of the society they live in: holding progressive attitudes in a traditional
society (or vice versa) may carry a well-being cost, while holding attitudes aligned
with the surrounding culture may not. This suggests any relationship between gender
attitudes and life satisfaction should not be assumed constant across countries, should be treated as an empirical question.

## Research Question

**RQ:** How do gender role attitudes relate to life satisfaction, and does this
relationship differ across countries?

**Hypotheses:**

-   **H1:** More progressive attitudes toward women's work will show a small
    average association with life satisfaction once demographic differences are
    accounted for.
-   **H2 (moderation):** The direction and/or strength of this association will
    vary by country, reflecting differing cultural norms around gender roles.

## Data

WVS Cross-National Wave 7 dataset (`WVS_Cross-National_Wave_7_csv_v6_0.csv`).

------------------------------------------------------------------------

# Data Setup

## Load packages

```{r load-packages}
# NOTE: MASS is loaded before tidyverse on purpose. MASS::select() would
# otherwise mask dplyr::select() and break every select() call below.
library(MASS)         # ordinal logistic regression (polr)
library(tidyverse)   # data wrangling + ggplot2 (dplyr::select overrides MASS::select)
library(gridExtra)   # combining plots
library(broom)        # tidy model output
library(estimatr)     # cluster-robust SEs (lm_robust)
library(car)          # VIF for multicollinearity checks
library(modelsummary) # side-by-side regression tables
```

## Import data

```{r import-data}
# Place WVS_Cross-National_Wave_7_csv_v6_0.csv in the same directory as this file.
# Using a relative path (rather than a hardcoded setwd()) keeps the project
# portable across machines.
wvs_data <- read_csv("WVS_Cross-National_Wave_7_csv_v6_0.csv")
```

## Data inspection

```{r inspect-data}
dim(wvs_data)
```

**How many observations / variables are in the raw dataset?** 97,220 observations
and 613 variables.

## Data cleaning

Variables used:

-   `B_COUNTRY_ALPHA` -- country
-   `S018` -- age (already scaled 0-1 in this wave)
-   `Q260` -- gender (1 = Male, 2 = Female)
-   `Q262` -- education
-   `Q46` -- life satisfaction (1 = dissatisfied ... 4 = satisfied, after recode)
-   `Q132` -- attitude toward women's participation in the workforce

```{r clean-data}
wvs_clean <- wvs_data %>%
  select(B_COUNTRY_ALPHA, S018, Q260, Q262, Q46, Q132) %>%
  rename(
    country = B_COUNTRY_ALPHA,
    age = S018,
    gender = Q260,
    education = Q262,
    life_satisfaction = Q46,
    women_work_attitude = Q132
  ) %>%
  filter(
    age > 0,
    gender %in% c(1, 2),
    education >= 1,
    life_satisfaction > 0,
    women_work_attitude > 0,
    country %in% c("USA", "GBR", "DEU", "JPN", "IND", "MEX")
  ) %>%
  # reverse-code so higher = more progressive (women's work seen as acceptable)
  mutate(women_work_attitude = 5 - women_work_attitude) %>%
  na.omit()

dim(wvs_clean)
```

**Q: How was the data cleaned, and how many observations/variables remain?**

The data were cleaned by selecting six relevant variables, dropping negative
(missing/refused) response codes, restricting the sample to six countries
(USA, UK, Germany, Japan, India, Mexico) for a tractable but still culturally
diverse cross-national comparison, and reverse-coding the work-attitude item so
higher values consistently mean "more progressive." The cleaned dataset contains
**11,093 observations and 6 variables**.

------------------------------------------------------------------------

# Analysis

## Descriptive statistics

```{r descriptive-statistics}
cat("NUMERIC VARIABLES\n")

cat("\nLife Satisfaction:\n")
print(wvs_clean %>% summarise(
  Mean = mean(life_satisfaction), Median = median(life_satisfaction),
  SD = sd(life_satisfaction), Min = min(life_satisfaction), Max = max(life_satisfaction)
))

cat("\nWomen's Work Attitude (higher = more progressive):\n")
print(wvs_clean %>% summarise(
  Mean = mean(women_work_attitude), Median = median(women_work_attitude),
  SD = sd(women_work_attitude), Min = min(women_work_attitude), Max = max(women_work_attitude)
))

cat("\nAge (scaled 0-1):\n")
print(wvs_clean %>% summarise(
  Mean = mean(age), Median = median(age), SD = sd(age), Min = min(age), Max = max(age)
))

cat("\nEducation:\n")
print(wvs_clean %>% summarise(
  Mean = mean(education), Median = median(education), SD = sd(education),
  Min = min(education), Max = max(education)
))

cat("\n\nCATEGORICAL VARIABLES\n")
cat("\nGender (1=Male, 2=Female):\n")
print(table(wvs_clean$gender))
print(prop.table(table(wvs_clean$gender)))

cat("\nCountry:\n")
print(table(wvs_clean$country))
print(prop.table(table(wvs_clean$country)))
```

```{r plot-descriptives, fig.height=5.5, fig.width=6.5}
plot1 <- ggplot(wvs_clean, aes(x = life_satisfaction)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Life Satisfaction", x = "Life Satisfaction", y = "Frequency") +
  theme_minimal()

plot2 <- ggplot(wvs_clean, aes(x = women_work_attitude)) +
  geom_histogram(binwidth = 0.5, fill = "pink", color = "black") +
  labs(title = "Distribution of Women's Work Attitude", x = "Women's Work Attitude", y = "Frequency") +
  theme_minimal()

plot3 <- ggplot(wvs_clean, aes(x = factor(gender, labels = c("Male", "Female")))) +
  geom_bar(fill = "purple", color = "black") +
  labs(title = "Gender Distribution", x = "Gender", y = "Frequency") +
  theme_minimal()

plot4 <- ggplot(wvs_clean, aes(x = country)) +
  geom_bar(fill = "darkgreen", color = "black") +
  labs(title = "Country Distribution", x = "Country", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

gridExtra::grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)
```

```{r correlation-heatmap, fig.height=4, fig.width=5}
# Correlation heatmap gives a quick visual overview before modeling
numeric_vars <- wvs_clean %>% select(age, education, life_satisfaction, women_work_attitude)
corr_mat <- cor(numeric_vars, use = "complete.obs")

corr_df <- as.data.frame(as.table(corr_mat))

ggplot(corr_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Freq, 2)), size = 3.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                        midpoint = 0, limits = c(-1, 1), name = "r") +
  theme_minimal() +
  labs(title = "Correlation Matrix: Numeric Variables", x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Descriptive summary:** Life satisfaction (mean = 1.77, median = 2, SD = 0.67) and
women's work attitude (mean = 1.99, median = 2, SD = 0.93) are both fairly centered
with meaningful spread. The sample is nearly evenly split by gender (49.6% male,
50.4% female) and unevenly spread across countries, ranging from Japan (11.0%) to
the USA (23.3%) -- a reminder that model results are not equally precise for every
country and that the pooled sample is not population-weighted.

## Modeling: does the relationship hold, and does it vary by country?

Four nested models test the research question directly:

1.  **Model 1** -- bivariate baseline
2.  **Model 2** -- adds gender and country as controls (main-effects only)
3.  **Model 3** -- adds a `women_work_attitude x country` interaction (tests H2)
4.  **Model 4** -- adds a `women_work_attitude x gender` interaction

```{r models}
model1 <- lm(life_satisfaction ~ women_work_attitude, data = wvs_clean)

model2 <- lm(life_satisfaction ~ women_work_attitude + factor(gender) + factor(country),
             data = wvs_clean)

model3 <- lm(life_satisfaction ~ women_work_attitude * factor(country) + factor(gender),
             data = wvs_clean)

model4 <- lm(life_satisfaction ~ women_work_attitude * factor(gender) + factor(country),
             data = wvs_clean)

modelsummary(
  list("Bivariate" = model1, "+ Controls" = model2,
       "x Country" = model3, "x Gender" = model4),
  stars = TRUE,
  gof_omit = "IC|Log|F"
)
```

**Test of moderation:** an F-test compares Model 2 (main effects only) to Model 3
(adds the country interaction) to formally test whether the attitude-satisfaction
slope differs by country.

```{r anova-test}
anova(model2, model3)
```

### Cluster-robust standard errors

Respondents are nested within six countries, so ordinary standard errors likely
understate uncertainty. As a robustness check, Model 2 is re-estimated with
standard errors clustered by country.

```{r robust-se}
model2_robust <- lm_robust(
  life_satisfaction ~ women_work_attitude + factor(gender) + factor(country),
  data = wvs_clean, clusters = country, se_type = "stata"
)
summary(model2_robust)
```

### Regression diagnostics

```{r diagnostics, fig.height=3.5, fig.width=6.5}
diag_df <- augment(model2)

p_resid <- ggplot(diag_df, aes(.fitted, .resid)) +
  geom_point(alpha = 0.15) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted", x = "Fitted values", y = "Residuals") +
  theme_minimal()

p_qq <- ggplot(diag_df, aes(sample = .resid)) +
  stat_qq(alpha = 0.15) +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot") +
  theme_minimal()

gridExtra::grid.arrange(p_resid, p_qq, ncol = 2)

cat("Variance Inflation Factors (Model 2):\n")
print(car::vif(model2))
```

Residuals show the expected banding of an ordinal 1-4 outcome fit with OLS,
and VIFs are all low (no meaningful multicollinearity among predictors). The
banding is exactly why an ordinal-regression robustness check (below) matters.

### Robustness check: ordinal logistic regression

Life satisfaction is genuinely ordinal (1-4), not continuous, so OLS is a
convenient approximation rather than the "correct" model. As a check, the same
relationship is estimated with a proportional-odds (ordinal logistic) model.

```{r ordinal-model}
wvs_clean_ord <- wvs_clean %>%
  mutate(life_satisfaction_f = factor(life_satisfaction, ordered = TRUE))

ord_model <- MASS::polr(
  life_satisfaction_f ~ women_work_attitude + factor(gender) + factor(country),
  data = wvs_clean_ord, Hess = TRUE
)

ord_summary <- coef(summary(ord_model))
p_vals <- pnorm(abs(ord_summary[, "t value"]), lower.tail = FALSE) * 2
ord_summary <- cbind(ord_summary, "p value" = round(p_vals, 4))
print(round(ord_summary, 3))
```

The sign and (approximate) magnitude of the `women_work_attitude` coefficient
match the OLS result, suggesting the linear-model conclusions are not an
artifact of treating an ordinal variable as continuous.

### Visualizing relationships

```{r plot-relations, fig.height=6, fig.width=6.5}
plot_scatter <- ggplot(wvs_clean, aes(x = women_work_attitude, y = life_satisfaction)) +
  geom_point(alpha = 0.3, color = "lightgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen", fill = "darkgreen", alpha = 0.15) +
  labs(title = "Life Satisfaction vs Women's Work Attitude",
       x = "Women's Work Attitude", y = "Life Satisfaction") +
  theme_minimal()

plot_gender <- ggplot(wvs_clean, aes(x = life_satisfaction,
                                     fill = factor(gender, labels = c("Male", "Female")))) +
  geom_histogram(binwidth = 0.5, alpha = 0.6, position = "dodge") +
  scale_fill_manual(values = c("Male" = "steelblue", "Female" = "red")) +
  labs(title = "Life Satisfaction by Gender", x = "Life Satisfaction", y = "Frequency", fill = "Gender") +
  theme_minimal()

# This faceted panel is the key visual test of H2: are the slopes different by country?
plot_country <- ggplot(wvs_clean, aes(x = women_work_attitude, y = life_satisfaction, color = country)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.6) +
  facet_wrap(~country) +
  labs(title = "Life Satisfaction vs Women's Work Attitude, by Country",
       subtitle = "Differing slopes indicate the relationship is moderated by country context",
       x = "Women's Work Attitude", y = "Life Satisfaction") +
  theme_minimal() + theme(legend.position = "none")

# Coefficient plot: interaction terms from Model 3
coef_df <- tidy(model3, conf.int = TRUE) %>%
  filter(str_detect(term, "women_work_attitude"))

plot_coef <- ggplot(coef_df, aes(x = reorder(term, estimate), y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Women's Work Attitude Slope, by Country (Model 3)",
       x = NULL, y = "Coefficient (95% CI)") +
  theme_minimal()

gridExtra::grid.arrange(plot_scatter, plot_gender, ncol = 2)
plot_country
plot_coef
```

## Interpretation

The pooled bivariate relationship between women's work attitude and life
satisfaction is weak and not statistically significant (Model 1), consistent
with the original lab's finding. Adding country and gender fixed effects (Model
2) doesn't change that conclusion. The interaction model (Model 3) and the
formal F-test comparing Models 2 and 3 test H2 directly: whether the slope
differs meaningfully by country, rather than relying on visual inspection
alone. The faceted plot above makes the same comparison visually -- some
countries show a near-flat or even negative slope while others show a mild
positive one, which is the pattern a "cultural fit" account of well-being
would predict, though the pooled cross-sectional design here cannot establish
causal direction.

------------------------------------------------------------------------

# Limitations

-   **Cross-sectional design**: no causal claims about attitudes -> satisfaction
    (reverse causality and confounding by unobserved country-level factors like
    GDP or welfare policy are both plausible).
-   **Non-representative sample**: unweighted WVS respondents from 6 countries,
    not a population-weighted comparison.
-   **Ordinal outcome modeled with OLS**: mitigated here with an ordinal logistic
    robustness check, but interaction effects were not re-tested in that
    specification due to sample-size-per-cell constraints.
-   **Single-item measures**: both life satisfaction and work attitudes are
    single survey items rather than validated multi-item scales.

# Conclusion

Gender role attitudes show, at most, a weak average association with life
satisfaction in this six-country sample, but the strength and direction of that
association appears to vary meaningfully by country -- supporting the idea that
country-level cultural context moderates how gender attitudes relate to
well-being, rather than there being one universal relationship.

------------------------------------------------------------------------

# Reproducibility

```{r session-info}
sessionInfo()
```
